SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2012-12-28 paper/paper.tex
paper: extrapolation section.

\todo{note constants are user-defined arguments}

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
aspect of the Haskell development cycle.  Because QuickCheck generates random
values for testing, counterexamples it finds may not be substantially larger
than a minimal counterexample.  In their original QuickCheck paper~\cite{qc},
the authors report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a]
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

@@

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
For example, consider the \ttp{shrink} instances given to data types \ttp{A} and
\ttp{B} in Figure~\ref{fig:initial}.\footnote{All examples and algorithms in
  this paper are in Haskell~\cite{haskell98}}.  Indeed, because of the regular
structure of \ttp{shrink} implementations, the principal motivation for one of
Haskell's first generics package was for providing generic definitions for
\ttp{shrink}~\cite{syb}.

However, this straightforward approach to defining \ttp{shrink} instances may
either miss minimal counterexamples, be too inefficient to be practical, or
both.\footnote{E.g., for a recent real-world case, see the user question posed
  on StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}
For example, consider again Figure~\ref{fig:initial}.  The data type \ttp{B} is
a product type over lists of 16-bit signed integers, wrapped by a constructor.
Consider the property \ttp{prop} in the figure, stating that if the sum of each
list in a field of \ttp{B} is less than 16, then the sum of the four fields of
\ttp{B} is less than 64.  The property \ttp{prop} seems reasonable at first
glance, until one realizes that due to underflow, the property can be violated.
For example, consider the value
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
Without shrinking, QuickCheck finds large examples to the property; on average,
QuickCheck returns a value contains 70 \ttp{Int16} values!

Thus, it pays to define \ttp{shrink}.  Unfortunately, even for a simple property
and program like this one, the definition for \ttp{shrink} given in
Figure~\ref{fig:initial} produces an intractable number of potential
counterexamples to test; with the definitions and property provided, which also
rely on QuickCheck's default instances for the list type, the list of potential
counterexamples generated by \ttp{shrink} can contain more than $10^{10}$
elements.  This is an enormous state space to search for a smaller example!  In
particular, if a counterexample is not found does not finish in a reasonable
amount of time.

It is not always straightforward to determine how to make a more efficient
implementation for \ttp{shrink}.  Shrinking presents a programmer's dilemma:
shrinking is used to better understand a counterexample to understand why a
property failed; but to implement an efficient shrink function in general
requires knowing how to generate counterexamples to the property.

@@

With SmallCheck, a property is checked up to a depth set by the user, measuring
how large values are.  For example, for type \ttp{Int}, the depth $d$ is defined
as the integers in the list enumerating from $-d$ to $d$ (i.e.,
\ttp{[(-d)..d]}).  For a product type, SmallCheck must check values produced by
taking the cross-product of the type's fields at a given depth.  In
Figure~\ref{fig:initial}, we have defined instances for generating Lazy
SmallCheck (a more efficient version of SmallCheck) tests by defining instances
of the \ttp{Serial} class.  For the property \ttp{prop} defined above (suitably
redefined to use SmallCheck's types), SmallCheck must check to depth 16 to find
the first counterexample.  Unfortunately, after a couple hours of testing, Lazy
SmallCheck is still checking values at depth six, and the number of tests scales
exponentially with respect to the depth.\footnote{All tests in this paper are
  performed on a four-core Pentium~i7 running at 2.7GHz with 8GB RAM on
  Fedora~16.  This test, and others unless noted, are performed using
  interpreted Haskell under GHC 7.4.2.}

@@

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Table~\ref{fig:results}.  The table characterizes testing the program shown
in Figure~\ref{fig:initial} that we have described so far.  Four trials are
presented.  In the first three, we use QuickCheck to generate counterexamples.
The first column contains the results of using QuickCheck without shrinking to
provide a baseline.  The next two columns show the result of using QuickCheck
with definitions of \ttp{shrink} that limit the lists of fields of \ttp{B} to 10
(Figure~\ref{lst:newshrink}) and 20 elements (Figure~\ref{lst:newshrink}, with
\ttp{len = 20}), respectively.  (Recall that with the original definition for
shrink from Figure~\ref{fig:initial}, some counterexamples cause QuickCheck to
not return after hours of computation.)  In each row, we show the maximum, mean,
and median number of seconds required by QuickCheck to generate a counterexample
over 100 runs.  In the last row, we show the mean size of the final value
returned, measured by the taking the cumulative length of the lists of the
fields of \ttp{B}.  Notice that while typically counterexamples are returned in
a fraction of a second, there are outliers taking 21 seconds and 125 seconds for
the two tests.  Both tests roughly cut the size of the counterexample in half;
increasing the number of potential counterexamples can result in a significant
performance penalty for this example without much benefit.

@@

\paragraph{Sum-type counterexample hypothesis}
First, recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the 0th sub-value.  The 0th sub-value of \ttp{cex} is
\ttp{cex} itself.  No invariant of the algorithm would be violated by beginning
with the 0th sub-value; recall that it would generate strictly smaller values
than \ttp{cex}.  However, we begin with a strict sub-value because of a
hypothesis we refer to as the \emph{sum-type counterexample hypothesis}.  The
hypothesis is that in the search space of possible values, if a counterexample
of the form \ttp{C}~$\vec{x}$, where \ttp{C} is some constructor and $\vec{x}$
are its arguments, is discovered, then it is more likely to find another
``nearby'' counterexample of the form \ttp{C}~$\vec{y}$, where $\vec{y}$ are
different arguments to \ttp{C}.  For example, consider a property about
unintended variable capture over a language's parse tree represented by a sum
type with constructors for module imports, function definitions, and
global-variable assignments, respectively.  A function definition counterexample
can only be reduced to smaller function definition counterexamples, the only
construct in which variable capture is possible.

@@

\paragraph{Conditional properties}
A QuickCheck property may be a conditional of the form
%
\begin{code}
prop a = pre a ==> prop' a
\end{code}
%
\noindent
such that \ttp{prop} is true if the precondition \ttp{pre a} is false or
\ttp{prop' a} holds.  In practice, conditionals are used to restrict the set of
values tested; values that fail the precondition are discarded.  Thus, the call
to \ttp{pass} in the reduction algorithm only considers a potential
counterexample \ttp{cex} to have failed if it passes the property's
precondition and fails the consequent.

\paragraph{Sub-value counterexample hypothesis}
Sometimes, a counterexample fails a property due to a sub-value nested deep
inside the counterexample.  The rest of the value is irrelevant.  We call this
the \emph{sub-value counterexample hypothesis.}  Thus, one way to efficiently
search the space of potential counterexamples is to test a counterexample's
(well-typed) sub-values.

For example, consider a simple calculator language containing constant
\ttp{Ints}, addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\noindent
and consider an interpreter for the language that returns \ttp{Nothing} if
division-by-zero is encountered and \ttp{Just} the value otherwise:
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Suppose we restrict the set of programs with the following predicate:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =    divSubTerms e0
                            && divSubTerms e1
divSubTerms (Div e0 e1)   =    divSubTerms e0
                            && divSubTerms e1
\end{code}
%
and then test the property
%
\begin{code}
div_prop e = divSubTerms e ==> eval e /= Nothing
\end{code}
%
\noindent
The property \ttp{div\_prop} turns out to be false; for example,
%
\begin{code}
Div (C 1) (Div (C 0) (C 1))
\end{code}
%
\noindent
is a counterexample.  In the predicate case
%
\begin{code}
divSubTerms (Div _ (C 0)) = False
\end{code}
%
\noindent
we forgot to recur on the divisor.

@@

\paragraph{Base types}

@@

\paragraph{Bounding counterexample exploration}

@@

\begin{figure}
  \begin{code}
extrapolate :: SubTypes a
       => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs =
    case index cex idx of
      Nothing -> return idxs
      Just v  -> do
        vs <- newVals v scMaxSize scMaxSuccess
        if allFail cex idx vs prop
          then extrapolate' (idx+1) (idx:idxs)
          else extrapolate' (idx+1) idxs

@@

The extrapolation algorithm is shown in Figure~\ref{fig:universal}.  The
algorithm is similar to the reduction algorithm in Figure~\ref{fig:reduction},
and in the implementation, the algorithms are generalized and combined.  The
function \ttp{extrapolate} returns a list of indexes to be generalized in the
original counterexample.  New sub-values are generated by \ttp{newVals}, shown
in Figure~\ref{fig:reduction}.  Here, we bound the size of values to generate by
the constant \ttp{scMaxSize}, which is independent of the size of the particular
sub-value.

@@

@@

>>>>

\todo{note constants are user-defined arguments, other simplifications}

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
aspect of the Haskell development cycle.  Because QuickCheck generates random
values for testing, counterexamples it finds may be substantially larger
than a minimal counterexample.  In their original QuickCheck paper~\cite{qc},
the authors report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a]
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

@@

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
For example, consider the \ttp{shrink} instances given to data types \ttp{A} and
\ttp{B} in Figure~\ref{fig:initial}.\footnote{All examples and algorithms in
  this paper are presented in Haskell~\cite{haskell98}.}  Indeed, because of the regular
structure of \ttp{shrink} implementations, the principal motivation for one of
Haskell's first generics package was for providing generic definitions for
\ttp{shrink}~\cite{syb}.

However, this straightforward approach to defining \ttp{shrink} instances may
either miss minimal counterexamples, be too inefficient to be practical, or
both.\footnote{For a recent real-world case, see the user question posed on
  StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}
For example, consider again Figure~\ref{fig:initial}.  The data type \ttp{B} is
a product type over lists of 16-bit signed integers, wrapped by a constructor.
Consider the property \ttp{prop} in the figure, stating that if the sum of each
list in a field of \ttp{B} is less than 16, then the sum of the four fields is
less than 64.  The property \ttp{prop} seems reasonable at first glance, until
one realizes that due to underflow, the property can be violated.  For example,
consider the value
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
Without shrinking, QuickCheck finds large examples to the property; on average,
QuickCheck returns a value contains 70 \ttp{Int16} values!

Thus, it pays to define \ttp{shrink}.  Unfortunately, even for a simple property
and program like this one, the definition for \ttp{shrink} given in
Figure~\ref{fig:initial} produces an intractable number of potential
counterexamples to test; with the definitions and property provided, which also
rely on QuickCheck's default instances for the list type, the list of potential
counterexamples generated by \ttp{shrink} can contain more than $10^{10}$
elements.  This is an enormous state space to search for a smaller example and
can cause the shrinking phase to take orders-of-magnitude more time than the
counterexample discovery phase.

It is not always straightforward to determine how to make a more efficient
implementation for \ttp{shrink}.  Shrinking presents a programmer's dilemma:
shrinking is used to better understand a counterexample to understand why a
property failed, but to implement an efficient shrink function in general
requires knowing how to generate counterexamples to the property.

@@

With SmallCheck, a property is checked up to a depth set by the user, measuring
how large values are.  For example, for type \ttp{Int}, the depth $d$ is defined
as the integers in the list enumerating from $-d$ to $d$ (i.e.,
\ttp{[(-d)..d]}).  For a product type, SmallCheck must check values produced by
taking the cross-product of the type's fields at a given depth.  In
Figure~\ref{fig:initial}, we have defined instances for generating Lazy
SmallCheck (a variant of SmallCheck) tests by defining instances
for the \ttp{Serial} class.  For the property \ttp{prop} defined above (suitably
redefined to use SmallCheck's types), SmallCheck must check to depth 16 to find
the first counterexample.  Unfortunately, after a couple hours of testing, Lazy
SmallCheck is still checking values at depth six, and the number of tests scales
exponentially with respect to the depth.\footnote{All tests in this paper are
  performed on a four-core Pentium~i7 running at 2.7GHz with 8GB RAM on
  Fedora~16.  This test, and others unless noted, are performed using
  interpreted Haskell under GHC 7.4.2.}

@@

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as we noted, it is infeasible for an example like this).  In the first
three columns, the results of using QuickCheck to generate counterexamples are
shown.  The first column contains the results of using QuickCheck without
shrinking to provide a baseline.  The next two columns show the result of using
QuickCheck with definitions of \ttp{shrink} that limit the lists of fields of
\ttp{B} to 10 (Figure~\ref{lst:newshrink}) and 20 elements
(Figure~\ref{lst:newshrink}, with \ttp{len = 20}), respectively.  (Recall that
with the original definition for shrink from Figure~\ref{fig:initial}, some
counterexamples cause QuickCheck to not return after hours of computation.)  In
each row, we show the maximum, mean, and median number of seconds required by
QuickCheck to generate a counterexample over 100 runs.  In the last row, we show
the mean size of the final value returned, measured by the taking the cumulative
length of the lists of the fields of \ttp{B}.  Notice that while typically
counterexamples are returned in a fraction of a second, there are outliers
taking 21 seconds and 125 seconds for the two tests.  Both tests roughly cut the
size of the counterexample in half; increasing the number of potential
counterexamples can result in a significant performance penalty for this example
without much benefit.

@@

\subsubsection{Sum-type counterexample hypothesis}
First, recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the 0th sub-value.  The 0th sub-value of \ttp{cex} is
\ttp{cex} itself.  No invariant of the algorithm would be violated by beginning
with the 0th sub-value; recall that it would generate strictly smaller values
than \ttp{cex}.  However, we begin with a strict sub-value because of a
hypothesis we refer to as the \emph{sum-type counterexample hypothesis}.  The
hypothesis is that in the search space of possible values, if a counterexample
of the form \ttp{C}~$\vec{x}$, where \ttp{C} is some constructor and $\vec{x}$
are its arguments, is discovered, then it is more likely to find another
``nearby'' counterexample of the form \ttp{C}~$\vec{y}$, where $\vec{y}$ are
different arguments to \ttp{C}.  For example, consider a property about
unintended variable capture over a language's parse tree represented by a sum
type with constructors for module imports, function definitions, and
global-variable assignments, respectively.  A function definition counterexample
can only be reduced to smaller function definition counterexamples, the only
construct in which variable capture is possible.

@@

\subsubsection{Conditional properties}
A QuickCheck property may be a conditional of the form
%
\begin{code}
prop a = pre a ==> prop' a
\end{code}
%
\noindent
such that \ttp{prop} is true if the precondition \ttp{pre a} is false or
\ttp{prop' a} holds.  In practice, conditionals are used to restrict the set of
values tested; values that fail the precondition are discarded.  Thus, the call
to \ttp{pass} in the reduction algorithm only considers a potential
counterexample \ttp{cex} to have failed if it passes the property's
precondition and fails the consequent.

\subsubsection{Sub-value counterexample hypothesis}\label{sec:subval}
Sometimes, a counterexample fails a property due to a sub-value nested deep
inside the counterexample.  The rest of the value is irrelevant.  We call this
the \emph{sub-value counterexample hypothesis.}  Thus, one way to efficiently
search the space of potential counterexamples is to test a counterexample's
(well-typed) sub-values.

For example, consider a simple calculator language containing constant
\ttp{Ints}, addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\noindent
and consider an interpreter for the language that returns \ttp{Nothing} if
division-by-zero is encountered and \ttp{Just} the value otherwise:
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Suppose we restrict the set of programs with the following predicate:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =    divSubTerms e0
                            && divSubTerms e1
divSubTerms (Div e0 e1)   =    divSubTerms e0
                            && divSubTerms e1
\end{code}
%
and then test the property
%
\begin{code}
prop_div e = divSubTerms e ==> eval e /= Nothing
\end{code}
%
\noindent
The property \ttp{prop\_div} turns out to be false; for example,
%
\begin{code}
Div (C 1) (Div (C 0) (C 1))
\end{code}
%
\noindent
is a counterexample.  In the predicate case
%
\begin{code}
divSubTerms (Div _ (C 0)) = False
\end{code}
%
\noindent
we forgot to recur on the divisor.

@@

\subsubsection{Base types}

@@

\subsubsection{Bounding counterexample exploration}

@@

\begin{figure}
  \begin{code}
extrapolate :: SubTypes a
       => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | or (map (subTree cex idx) idxs) =
      extrapolate' (idx+1) idxs
  extrapolate' idx idxs
    | otherwise                       =
      case index cex idx of
        Nothing -> return idxs
        Just v  -> do
          vs <- newVals v scMaxSize scMaxSuccess
          if allFail cex idx vs prop
            then extrapolate' (idx+1) (idx:idxs)
            else extrapolate' (idx+1) idxs

@@

The extrapolation algorithm is shown in Figure~\ref{fig:universal}.  The
algorithm is similar to the reduction algorithm in Figure~\ref{fig:reduction},
and in the implementation, the algorithms are generalized and combined.  The
function \ttp{extrapolate} returns a list of indexes to be generalized in the
original counterexample.  In the recursive function \ttp{extrapolate'}, there is
a function guard in which a call to \ttp{subTree}
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\ttp{subTree cex idx0 idx1} is true if in \ttp{cex}, the value at index
\ttp{idx0} is a child of index \ttp{idx1}, in a tree representation of the
data.  This prevents the algorithm from trying to generalize sub-values that are
abstracted away already since their parents have been generalized.

New sub-values are generated by \ttp{newVals}, shown in
Figure~\ref{fig:reduction}.  Here, we bound the size of values to generate by
the constant \ttp{scMaxSize}, which is independent of the size of the particular
sub-value.

@@

\todo{Add discussion here}

@@

Sum types denote choice in a data type.  Sometimes, a property over a sum type
fails because there is a bug for some of the variants but not others.  For
example, recall the calculator language from Section~\ref{sec:subval}.  The
property \ttp{prop\_div} fails only for values that contain a variant tagged
with the \ttp{Div} constructor.  For example, consider the following counterexample:
%
\begin{code}
Div (Add (C 20) (C 30))
    (Add (C 89) (C -89))
\end{code}
%
\noindent
The counterexample fails because the divsor evaluates to zero.  Using universal
sub-value generalization described above, the counterexample is generalized as
follows:
%
\begin{code}
forall x . Div x (Add (C 89) (C -89))
\end{code}
%
\noindent
So we know there is something particular about the \ttp{Add (C 89) (C -89)},
because there are some values that we can replace this sub-value with in the
original counterexample causing it to satisfy the property.  But we might wonder
if there is something special about variants tagged by the \ttp{Add}
constructor, or might we finding failing sub-values with the other variants
tagged by the constructors \ttp{C} and \ttp{Div}?  If so, then we know there is
not some bug being triggered by \ttp{Add} variants specifically, so it is more
likely that the bug involves how the second field of the \ttp{Div} variant is
evaluated.

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there a counterexample
containing every possible variant as a sub-value, then we abstract it.  For
example, the following all characterize counterexamples:
%
\begin{code}
forall x . Div x (Add (C 89) (C -89))
forall x . Div x (C 0)
forall x . Div x (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the second field of the outermost \ttp{Div}
constructor.  We therefore return a formula like the following:
%
\begin{code}
forall values x .
  forall constructors C\(\sb{0}\) .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (C\(\sb{0} \stackrel{\rightarrow}{y}\))
\end{code}
%
